\documentclass{article}

\author{Alexander Pluska}

\usepackage{url}
\usepackage{bussproofs}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{wasysym}
\usepackage{tikz}
\usetikzlibrary{cd}

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\theoremstyle{definition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{definition}
\newtheorem{remark}[theorem]{Remark}

\title{Dissertation Agreement - Exposé}

\newcommand{\A}{\mathbf A}
\newcommand{\B}{\mathbf B}
\renewcommand{\lim}{\mathbf{lim\:}}
\newcommand{\colim}{\mathbf{colim\:}}
\newcommand{\0}{\mathbf 0}
\newcommand{\1}{\mathbf 1}
\newcommand{\id}{\text{id}}
\newcommand{\eq}{\text{eq}}
\newcommand{\coeq}{\text{coeq}}
\newcommand{\Set}{\textbf{Set}}
\newcommand{\Sketch}{\textbf{Sketch}}
\newcommand{\Mod}{\textbf{Mod}}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcounter{question}
\newenvironment{question}{\smallskip\noindent\textbf{Question \refstepcounter{question}\arabic{question}}\begin{em}}{\end{em}}
\newenvironment{method}{\smallskip\noindent\textbf{Methodology}}{}

\begin{document}
	

	\maketitle

	\section{Introduction}
	
	Building on previous work~\cite{thesis,pluska2023embedding} the goal of my dissertation is to further the state of multiple aspects of theorem proving for intuitionistic logic.

	Intuitionistic logic refers to a flavor of logic in which the existence of an object can only be established by explicit construction, as opposed to classical logic where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction.

	It essentially differentiates itself from classical logic by the fact that the law of excluded middle $A\vee\neg A$ and the double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.
	Besides philosophical considerations, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation for studying intuitionistic logic from the perspective of computer science in that proofs directly correspond to computer programs --- as expressed in the Curry--Howard correspondence~\cite{howard1980formulae}. Importantly, most modern proof assistants such as Coq~\cite{bertot2013interactive} or Lean~\cite{de2015lean} are intuitionistic.

	The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems and a collection of benchmark problems (see e.g. the ILTP library website~\cite{iltp}).
	The progress in automated reasoning for intuitionistic logic, however, has been slower than the  impressive advances in solvers for classic logics --- evidenced, e.g., by the CASC~\cite{casc} and SAT~\cite{satc} competitions..
	This difference can partially be explained by fundamental differences between the logics.
	First of all, determining intuitionistic validity is computationally harder, e.g., in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic}, whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
	A further advantage of classical logic is the existence of calculi that are particularly suited for automation, such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability (i.e., in order to show the validity of a formula it suffices to show the unsatisfiability of the negated formula, which is insufficient in intuitionistic logic).
	The first dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient,tammet1996resolution} used the naïve inverse method, i.e., a direct search for a cut-free proof by applying the rules from some proof calculus inversely, which generally leads to a very complex search.
	More recently, connection-based methods have been applied to various non-classical logics~\cite{otten2005clausal,otten2021nanocop}, including intuitionistic logic.
	There have also been some successful attempts to study intuitionistic validity via embedding into higher-order classical logic~\cite{LEO}.
	However, in comparison to classical provers, intuitionistic provers are much less mature and many modern techniques developed for classical provers have not been tested in an intuitionistic setting.

	Building on my master's thesis~\cite{thesis} and previous paper~\cite{pluska2023embedding} the goal of my dissertation is to advance the state of theorem proving for intuitionistic logic and transfer known techniques and applications from theorem proving for classical logic. In particular I will focus on the following questions:
	\begin{itemize}
		\item Can techniques from state-of-the-art intuitionistic theorem provers~\cite{otten2008leancop,otten2021nanocop} and classical theorem provers~\cite{kovacs2013first,schulz2002brainiac} be combined to obtain a more efficient prover for intuitionistic logic? Current state-of-the-art intuitionistic provers feature a lean design~\cite{otten2008leancop,otten2021nanocop}, whereas state-of-the-art classical provers are generally much more complex~\cite{kovacs2013first}. This opens two possible avenues for merging the two approaches: either we can try to extend the complex classical provers with intuitionistic techniques to handle intuitionistic validity, or we can hand-pick a subset of the classical techniques that are particularly suited for intuitionistic logic and implement them in an intuitionistic prover. In particular modern approaches based on machine learning~\cite{kaliszyk2018reinforcement,loos2017deep,rawson2019neurally} could benefit from a lean underlying prover in terms of explainability and interpretability.
		\item A known challenge for current provers is finding counter-models to validity. Since even fewer statements are valid in intuitionistic logic, this problem is even more pronounced. Can we find a way to efficiently find counter-models to intuitionistic validity? We shall primarily focus on formulas that are not intuitionistically valid, but classically valid. However, the general problem of finding counter-models to validity is also of interest.
		\item One major application of classical theorem provers is automation of proof assistants via hammers~\cite{bohme2010sledgehammer}. While a hammer for constructive proof assistants exists~\cite{czajka2018hammer} it uses classical theorem provers and cannot generally utilize the generated proof. Can we develop a hammer for constructive proof assistants that utilizes an intuitionistic prover and its proofs?
	\end{itemize}

	\section{Background}

	\subsection{Automated theorem proving}

	\subsection{Intuitionistic Logic}

	\section{Research plan}

	\subsection{Problems}

	\begin{question}
		Can a syntactic embedding of intuitionistic logic into classical logic be leveraged to efficiently decide intuitionistic validity using a classical prover? Can we can we retrieve intuitionistic proofs/counter-models from the corresponding classical proofs/counter-models?
	\end{question}

	\vspace*{12pt}

	\begin{method}
		Utilizing the Kripke semantics for intuitionistic logic we devise a dual embedding to the famous embedding of classical logic into intuitionistic logic by G\"odel~\cite{goedel1933consistency}, implement this embedding to work on a standard problem format like TPTP~\cite{tptp} and apply it to a benchmark set of intuitionistic problems, e.g.~\cite{iltp}. We then apply a state-of-the-art classical prover to the embedded problems and compare the performance to state-of-the-art intuitionistic provers. This question has been partially adressed in my master's thesis~\cite{thesis} and during the first year, resulting in a paper~\cite{pluska2023embedding}.
	\end{method}

	\vspace*{12pt}

	\begin{question}
		Can a more deep reflection of intuitionistic semantics in the proof calculus be used to leverage the power of modern classical provers for intuitionistic logic?
	\end{question}

	\begin{method}
		It is possible to embed intuitionistic semantics more deeply into a proof calculus than as a syntactic transformation of the formula. This has already been done in connection calculi~\cite{otten2005clausal}. We will modify the resolution calculus~\cite{bachmair2001resolution} to reflect intuitionistic semantics and implement this calculus in a existing prover, e.g. vampire~\cite{kovacs2013first}. We will then compare the performance of this prover to state-of-the-art intuitionistic provers. Furthermore, we will experiment with other calculi and prover architectures.
	\end{method}

	\vspace*{12pt}

	\begin{question}
		Can we adapt modern techniques from classical provers to intuitionistic provers, in particular machine learning techniques?
	\end{question}

	\begin{method}
		State-of-the-art intuitionistic proversusch an leanCoP~\cite{otten2008leancop} and nanoCoP~\cite{otten2021nanocop} are generally much simpler than state-of-the-art classical provers such as Vampire~\cite{kovacs2013first}, iProver~\cite{korovin2008iprover} or E~\cite{schulz2002brainiac}. On the one hand modern classical provers augment their calculus to utilize different forms of redundacy to minimize the search space, see e.g.~\cite{bbachmair1994rewrite,hillenbrand2013search, gleiss2020subsumption} and on the other hand they utilize powerful portifolio modes~\cite{rawson2018dynamic} in order to optimize their large suites of hyperparameters. More recently, machine learning techniques have been applied to direct search either by predicting hyperparameters, e.g.~\cite{bartek2020learning,bartek2023much}, or replacing the usual ordering-based search, e.g.~\cite{jakubuuv2017enigma,kaliszyk2018reinforcement,crouse2021deep}. We will handpick a subset of these techniques that are particularly suited for intuitionistic logic and implement them in a state-of-the-art intuitionistic prover, e.g. nanoCoP~\cite{otten2021nanocop}. We will then compare the performance of this prover to state-of-the-art intuitionistic provers. In particular in case of machine learning techniques we expect benefits in terms of explainability and interpretability due to the lean underlying prover.
	\end{method}

	\vspace*{12pt}

	\begin{question}
		Can we replace classical provers in settings where intuitionistic provers would be more suitable?
	\end{question}

	\begin{method}
		Sometimes it is desirable to obtain a intuitionistic proof for a problem rather than a classical one, e.g. in the context of proof assistants based on constructive type theory such as Lean~\cite{de2015lean} or Coq~\cite{bertot2013interactive} or in program synthesis. There have been attempts to use classical provers in these context, e.g.~\cite{czajka2018hammer,hozzova2023program}. However, these approaches are limited in that they cannot generally utilize the generated proof and a successful proof does not even guarantee the existence of a solution. We will investigate whether we can replace classical provers in these settings with intuitionistic provers. 
	\end{method}

	\subsection{Timeline}

	\subsection{Lectures and conferences}
	


	\bibliographystyle{acm}
	\bibliography{bibliography}
	
\end{document}